---
title: "gastric-cancer-survival-analysis"
format: html
editor: visual
---

# Gastric cancer survival analysis

Gastric cancer is one of the leading causes of cancer-related mortality worldwide, with overall survival strongly influenced by tumour stage at diagnosis. While TNM staging provides essential prognostic information, patients within the same stage group show considerable variation in outcomes, suggesting that molecular factors not captured by pathological staging play an important role. This analysis uses genome-wide gene expression data from 433 primary gastric cancer patients (GSE84437) to identify genes whose expression is associated with overall survival independently of age, T stage, and N stage. By combining Cox proportional hazards regression with false discovery rate correction, we aim to identify candidate prognostic biomarkers that could complement clinical staging and potentially guide treatment decisions.

```{r}
# Only need to run this once
# install.packages("survminer")   # for beautiful Kaplan-Meier plots
# install.packages("dplyr")       # for data wrangling
# install.packages("ggplot2")     # for plotting
# install.packages("dplyr")
# install.packages("knitr")
# install.packages("forcats")
```

```{r}
library(survival)
library(dplyr)
library(ggplot2)
library(GEOquery)
library(forcats)
library(survminer)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

## Dataset

GSE84437 includes gene expression profile data from 433 patients with primary gastric cancer. Alongside the gene expression data, it also contains detailed clinical characteristics such as gender, age, T stage, N stage, survival time, and survival status.

```{r}
gse_dir <- "/Users/robert/Documents/private_projects/gastro/data/GEO"
dir.create(gse_dir, recursive = TRUE, showWarnings = FALSE)
gse <- getGEO("GSE84437", GSEMatrix = TRUE, destdir = gse_dir)
```

Let us load the data:

```{r}
eset <- gse[[1]]         # ExpressionSet object
expr   <- exprs(eset)    # matrix: rows = probes/genes, cols = patients 
pheno  <- pData(eset)    # data.frame: rows = patients, cols = clinical variables
features <- fData(eset)  # data.frame: probe ID to gene symbol mapping

dim(expr)      # e.g., 47323 probes × 433 patients
head(pheno)    # see what clinical columns are available
```

```{r}
head(features)
```

We have:

-   **`expr`** (49,576 × 433)

    -   Rows = probes/genes

    -   Columns = patients

    -   Values = gene expression levels (numbers)

-   **`pheno`** (433 × \~40)

    -   Rows = patients

    -   Columns = clinical variables (age, sex, survival time, death, stage...)

    -   The patient IDs link `pheno` rows ↔ `expr` columns

-   **`features`** (49,576 × 30)

    -   Rows = probes

    -   Columns = probe annotations (ID, gene symbol, chromosome...)

    -   The probe IDs link `features` rows ↔ `expr` rows

A **probe** is a short DNA sequence on the microarray chip designed to detect a specific gene.

-   **One gene can have multiple probes** — the same gene might be measured by 2-5 different probes on the chip, as a redundancy measure.

-   **Some probes don't map to any gene** — they may be control probes or outdated sequences with no known gene symbol

-   **Some probes are ambiguous** — they might match multiple genes

```{r}
# We do not want probes but genes
# 1. Rename probes to gene symbols
rownames(expr) <- features$Symbol[match(rownames(expr), features$ID)]

# 2. Remove probes with no gene symbol
dim(expr)
expr <- expr[!is.na(rownames(expr)), ]
dim(expr)

# 3. Collapse: keep highest-expressed probe per gene
expr <- expr[order(rowMeans(expr, na.rm = TRUE), decreasing = TRUE), ]
expr <- expr[!duplicated(rownames(expr)), ]
dim(expr)
```

```{r}
colnames(pheno)
```

```{r}
pheno <- pheno %>%
  mutate(
    time   = as.numeric(`duration overall survival:ch1`),
    event  = as.numeric(`death:ch1`),
    age    = as.numeric(`age:ch1`),
    sex    = as.factor(`Sex:ch1`),
    tstage = as.factor(`ptstage:ch1`),
    nstage = as.factor(`pnstage:ch1`)
  )

# Filter out nan entries
dim(pheno)
pheno <- pheno[!is.na(pheno$time) & !is.na(pheno$event), ]
dim(pheno)

# We also filter out the corresponding expr values
expr  <- expr[, rownames(pheno)]

dim(pheno)
```

Let us look at the columns essential for our analysis:

## `death:ch1` — the event indicator

This is what Cox regression calls the **event status** or **censoring indicator**. It tells you:

-   `1` = the patient **died** during the study follow-up period

-   `0` = the patient was **still alive** when the study ended (called **censored**)

Censored means: we don't know when or if they die, we just know they were alive up to a certain point. Cox regression handles this elegantly — it doesn't throw those patients away, it uses the information "survived at least this long."

## `duration overall survival:ch1` — the time variable

This is the **survival time** — how long the patient was followed from diagnosis until either:

-   They **died** (if `death = 1`), or

-   The **study ended / they were lost to follow-up** (if `death = 0`)

It's typically measured in **months** or **days**.

# Covariates

Before we look at the influence of the genes we will check the impact of the following covariates:

-   sex: gender of the patient

-   age

-   tstage: tumour depth (T1–T4)

-   nstage: lymph node involvement (N0–N3)

```{r}
hist(pheno$age,
     main  = "Age distribution",
     xlab  = "Age (years)",
     col   = "steelblue",
     border = "white",
     breaks = 20)
```

```{r}
sex_counts <- table(pheno$sex)

sex_df <- as.data.frame(sex_counts)
colnames(sex_df) <- c("sex", "count")

ggplot(sex_df, aes(x = "", y = count, fill = sex)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Sex distribution") +
  theme_void()

```

```{r}
ggplot(pheno, aes(x = tstage)) +
  geom_bar(fill = "steelblue", color = "white") +
  labs(title = "T stage distribution", x = "T stage", y = "Count")
```

```{r}
table(pheno$tstage)
```

Due to the low numbers in T1 and T2 we group them into early stage:

```{r}
pheno$tstage_collapsed <- as.factor(fct_collapse(pheno$tstage,
  "T1_T2" = c("T1", "T2"),
  "T3"    = "T3",
  "T4"    = "T4"
))
```

```{r}
ggplot(pheno, aes(x = nstage)) +
  geom_bar(fill = "steelblue", color = "white") +
  labs(title = "N stage distribution", x = "N stage", y = "Count")

```

```{r}
# Counts for each n stage
table(pheno$nstage)
```

```{r}
surv_object <- Surv(time  = pheno$time, event = pheno$event)

cox_model <- coxph(surv_object ~ age + sex + tstage_collapsed + nstage, data = pheno)
summary(cox_model)
```

According to this model we have:

**Significant:**

-   **Age** (p = 0.0004): each additional year increases hazard by \~2.3%.

-   **N2**: HR = 2.55 (p \< 0.001)

-   **N3**: HR = 3.10 (p \< 0.001)

-   **N1**: still borderline (p = 0.093)

-   **T3**: HR = 2.14 (p = 0.042) — 2× higher risk than early stage

-   **T4**: HR = 3.27 (p = 0.0007) — 3.3× higher risk, very convincing

**Not significant:**

-   **Sex**: no meaningful effect (HR = 1.19, wide CI, p = 0.26) — consistent with literature

-   **T2**: no meaningful effect

```{r}
ggcoxdiagnostics(cox_model, type = "dfbeta")
```

```{r}
test_ph <- cox.zph(cox_model)
print(test_ph)
```

We observe that the PH assumption is actually violated for both **T stage** (p=0.023) and **N stage** (p=0.047), and the global test confirms it (p=0.038). Age and sex are fine. The hazard ratio for stage is not constant over time. Early-stage patients may have similar short-term survival but diverge later, or high-stage patients have excess early mortality that diminishes over time.

Therefore, we consider a model where we stratify over tstage and nstage. That is we assume a different baseline risk for different t stages and n stages.

```{r}
cox_model <- coxph(surv_object ~ age + sex + strata(tstage_collapsed) + strata(nstage), data = pheno)
summary(cox_model)
```

**Age and sex estimates are stable:**

-   Age HR = 1.022, virtually identical to before (was 1.023) — reassuring

-   Sex still non-significant (p = 0.303) — consistent

Let us also look again at the covariates nstage and tstage using Kaplan Maier curves to confirm that are significant.

```{r}
km_fit <- survfit(surv_object ~ nstage, data = pheno)

ggsurvplot(
  km_fit,
  data          = pheno,
  pval          = TRUE,        # log-rank p-value
  conf.int      = TRUE,        # 95% CI ribbon
  risk.table    = TRUE,        # number at risk below plot
  legend.labs   = levels(pheno$nstage),
  xlab          = "Time (months)",
  ylab          = "Survival probability",
  title         = "Kaplan-Meier Survival Curves by nstage"
)
```

```{r}
km_fit <- survfit(surv_object ~ tstage, data = pheno)

ggsurvplot(
  km_fit,
  data          = pheno,
  pval          = TRUE,        # log-rank p-value
  conf.int      = TRUE,        # 95% CI ribbon
  risk.table    = TRUE,        # number at risk below plot
  legend.labs   = levels(pheno$tstage),
  xlab          = "Time (months)",
  ylab          = "Survival probability",
  title         = "Kaplan-Meier Survival Curves by tstage"
)
```

## Gene Expression Data

```{r}
# Let us check the metric of gene expression
pheno$data_processing[1]
```

```{r}
sum(is.na(pheno$time))
sum(is.na(pheno$event))
```

```{r}
# Some genes have no expression data
dim(expr)
expr <- expr[rowSums(!is.na(expr)) > 0, ]
dim(expr)

# Remove any rows with no symbol or empty string
dim(expr)
expr <- expr[!is.na(rownames(expr)) & rownames(expr) != "", ]
dim(expr)
```

```{r}
# Are there duplicates row names?
sum(duplicated(rownames(expr)))
```

```{r}
# We transfrom the gene expression to the log scale
expr_log <- log2(expr + 1)
```

```{r}
cox_pval <- function(gene_exp) {
  # gene_exp: A vector of 433 expression values, one per patient
  # Adds the expression values as a temporary column gene in pheno
  # Now pheno has a column linking each patient to their expression level for this gene
  pheno$gene <- gene_exp
  tryCatch({
    fit <- coxph(surv_object ~ gene + age + strata(tstage_collapsed) + strata(nstage), data = pheno)
    summary(fit)$coefficients[1, 5]
  }, error = function(e) NA)
}

results <- apply(expr_log, 1, cox_pval)

head(results)
```

Before we continue let us check if the PH assumptions also hold for the genetic markers:

```{r}
gene_of_interest <- "LOXL4"  # replace with any gene of choice
pheno$gene <- as.numeric(expr_log[gene_of_interest, ])
fit <- coxph(surv_object ~ gene + age + strata(tstage_collapsed) + strata(nstage), data = pheno)
test_ph <- cox.zph(fit)
plot(test_ph)  # For a visual check. This should resemble a flat line around 0
```

```{r}
print(test_ph)
```

```{r}
# FDR correction and filter
results_df <- data.frame(
  symbol  = names(results),
  pval    = results,
  fdr     = p.adjust(results, method = "BH")
)

# Let us see how many significant genes we found
sig_genes <- results_df[results_df$fdr < 0.05, ]
sig_genes <- sig_genes[order(sig_genes$fdr), ]
nrow(sig_genes)
```

Only one gene is significantly asssociated with a different survival rate.

```{r}
head(sig_genes)
```

This gene associates with survival **independently of age, T stage, and N stage**. Before we look at the effect size let us check the distribution of the gene expression of LOXL4.

```{r}
pheno$LOXL4   <- as.numeric(expr_log["LOXL4", ])
ggplot(pheno, aes(x = LOXL4 )) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = median(pheno$LOXL4), 
             color = "red", linetype = "dashed", linewidth = 0.8) +
  labs(title = "LOXL4 expression distribution",
       x     = "LOXL4 expression",
       y     = "Count")

```

```{r}
summary(as.numeric(expr_log["LOXL4", ]))
```

Now let us look at the effect size. We rescale the expression levels to obtain the effect size in terms of standard deviations of gene expressions.

```{r}
# We rescale from gene expression to standard deviation after log transformation. 
# => Each HR represents the effect of a 1 standard deviation increase in expression
pheno$LOXL4   <- scale(as.numeric(expr_log["LOXL4", ]))
fit <- coxph(surv_object ~ LOXL4 + age + strata(tstage_collapsed) + strata(nstage), data = pheno)

summary(fit)
```

Let us look at samples that influence the result overly.

```{r}
ggcoxdiagnostics(fit, type = "dfbeta")
```

```{r}
mart_res <- residuals(fit, type = "martingale")

plot(pheno$LOXL4, mart_res,
     xlab = "LOXL4 expression (log2, scaled)",
     ylab = "Martingale residuals",
     main = "Linearity check for LOXL4")
abline(h = 0, col = "red", lty = 2)
lines(lowess(pheno$LOXL4, mart_res), col = "blue")
```

We identify the outliers.

```{r}
# Extract deviance residuals manually
dev_res <- residuals(fit, type = "deviance")

# Flag large residuals (rule of thumb: |deviance| > 2)
outliers <- which(abs(dev_res) > 2)
length(outliers)
pheno[outliers, c("time", "event", "age", "tstage_collapsed", "nstage")]

```

Let us rerun the model and run it on the data without the outliers to assure the robustness of the result.

```{r}
# Remove outliers
pheno$LOXL4   <- scale(as.numeric(expr_log["LOXL4", ]))
pheno_clean <- pheno[!rownames(pheno) %in% names(outliers), ]

# Refit on cleaned data

fit_clean <- coxph(Surv(time, event) ~ LOXL4 + age + 
                   strata(tstage_collapsed) + strata(nstage), 
                   data = pheno_clean)

summary(fit_clean)
```

This confirms that the result was not driven by the outliers.

For illustration let us look at one Kaplan-Meier curve for LOXL4.

```{r}
pheno$LOXL4_group <- ifelse(pheno$LOXL4 > median(pheno$LOXL4), "High", "Low")
pheno$LOXL4_group <- as.factor(pheno$LOXL4_group)

km_fit_LOXL4 <- survfit(surv_object ~ LOXL4_group, data = pheno)

ggsurvplot(
  km_fit_LOXL4,
  data       = pheno,
  pval       = TRUE,
  conf.int   = TRUE,
  risk.table = TRUE,
  legend.labs = c("High", "Low"),
  xlab       = "Time (months)",
  ylab       = "Survival probability",
  title      = "Kaplan-Meier Survival Curves by LOXL4 expression"
)
```

## Biological interpretation

We find the following in the literature:

LOXL4 (Lysyl Oxidase Like 4) encodes an extracellular matrix enzyme involved in collagen and elastin crosslinking, which promotes tumour stiffening and facilitates invasion and metastasis. Elevated LOXL4 expression has been previously implicated in tumour progression across multiple cancer types, and its association with worse survival in this cohort — independently of clinical stage — suggests it may reflect an aggressive tumour microenvironment biology not captured by conventional staging.

## Outlook

These findings are exploratory in nature and require validation in an independent gastric cancer cohort before any clinical conclusions can be drawn.
