---
title: "gastric-cancer-survival-analysis"
format: html
editor: visual
---

```{r}
# Only need to run this once
# install.packages("survminer")   # for beautiful Kaplan-Meier plots
# install.packages("dplyr")       # for data wrangling
# install.packages("ggplot2")     # for plotting
# install.packages("dplyr")
# install.packages("knitr")
```

```{r}
library(survival)
library(dplyr)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

## Dataset

GSE84437 includes gene expression profile data from 433 patients with primary gastric cancer. Alongside the gene expression data, it also contains detailed clinical characteristics such as gender, age, T stage, N stage, survival time, and survival status.

```{r}
library(GEOquery)
gse <- getGEO("GSE84437", GSEMatrix = TRUE)
```

Let us load the data:

```{r}
eset <- gse[[1]]         # ExpressionSet object
expr   <- exprs(eset)    # matrix: rows = probes/genes, cols = patients 
pheno  <- pData(eset)    # data.frame: rows = patients, cols = clinical variables
features <- fData(eset)  # data.frame: probe ID to gene symbol mapping

dim(expr)      # e.g., 47323 probes × 433 patients
head(pheno)    # see what clinical columns are available
```

```{r}
colnames(pheno)
```

Let us look at the columns essential for our analysis:

## `death:ch1` — the event indicator

This is what Cox regression calls the **event status** or **censoring indicator**. It tells you:

-   `1` = the patient **died** during the study follow-up period

-   `0` = the patient was **still alive** when the study ended (called **censored**)

Censored means: we don't know when or if they die, we just know they were alive up to a certain point. Cox regression handles this elegantly — it doesn't throw those patients away, it uses the information "survived at least this long."

## `duration overall survival:ch1` — the time variable

This is the **survival time** — how long the patient was followed from diagnosis until either:

-   They **died** (if `death = 1`), or

-   The **study ended / they were lost to follow-up** (if `death = 0`)

It's typically measured in **months** or **days**.

```{r}
pheno$time  <- as.numeric(pheno$`duration overall survival:ch1`)
pheno$event <- as.numeric(pheno$`death:ch1`)
```

```{r}
surv_object <- Surv(time  = pheno$time, event = pheno$event)
```

```{r}
pheno <- pheno %>%
  mutate(
    time   = as.numeric(`duration overall survival:ch1`),
    event  = as.numeric(`death:ch1`),
    age    = as.numeric(`age:ch1`),
    sex    = as.factor(`Sex:ch1`),
    tstage = as.factor(`ptstage:ch1`),
    nstage = as.factor(`pnstage:ch1`)
  )
```

```{r}
head(pheno$`age`)
head(pheno$`sex`)
```

```{r}
cox_model <- coxph(surv_object ~ age + sex, data = pheno)
```

```{r}
summary(cox_model)
```
